Is high social class always beneficial for survival? A study of Northern Sweden

### 2017-11-11 article3.Rmd
Small revisions

We must however consider other and potentially more important explanations as well and that the embodied body can be materialized in different ways. It is possible that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantagous position. Such stress can also be expressed in unhealthful ways of living. These behaviours can however also have to do with certain expectations on being a man of high status. Our results of the gender aspect is important in this respect. As several researchers have pointed out, demographic research has often ignored power relations and different roles within families [@watkins93; @mack95; @jans07]. The gender difference in mortality may be caused by discrimination on the female part in the household. Historical Sweden was strongly gender segregated, discriminating women and giving them less space for action. In our study, we do not find the lack of power to have negative health consequences for women. Within the same households, the effect of social class leads to different results, high social position being beneficial for women but not for men. Instead, we might consider the role of behaviours based on distinct gendered and class expectations. Having a high social status is sometimes expressed in attitudes, norms and way of living potentially harmful for health. @edli11, finding similar results in another environment, suggest that an explanation for the larger social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking, smoking and more inclined to take risks.  Alcohol has been suggested as part of the large gender differences in the 19th century [@wil99; @edv92], and a more risk-taking life may be an important component of male mortality [@cour00]. Different historical contexts promote different ways to spend economic resources [@rasp06; @jaadla17].

@@

Overweight, obesity and physical inactivity were probably more common in higher social classes. @vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers, something they suggest can be explained by the higher tobacco consumption among men in these groups. The social pattern among women was however reversed in line with the present study. There is no solid information on smoking in different social classes before the 1960s, but there are international studies indicating higher prevalence and consumption of tobacco in higher classes during the early phases of the tobacco epidemic, illustrated for example by the fact that smoking initially was more prevalent among physicians. There was also an association between higher incomes and tobacco consumption [@loco94].

>>>>

We must however consider other and potentially more important explanations as well and that the embodied body can be materialized in different ways. It is possible that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantagous position. Such stress can also be expressed in unhealthful ways of living. These behaviours can however also have to do with certain expectations on being a man of high status. Our results of the gender aspect is important in this respect. As several researchers have pointed out, demographic research has often ignored power relations and different roles within families [@watkins93; @mack95; @jans07]. The gender difference in mortality may be caused by discrimination on the female part in the household. Historical Sweden was strongly gender segregated, discriminating women and giving them less space for action. In our study, we do not find the lack of power to have negative health consequences for women. Within the same households, the effect of social class leads to different results, high social position being beneficial for women but not for men. Instead, we might consider the role of behaviours based on distinct gendered and class expectations. Having a high social status is sometimes expressed in attitudes, norms and way of living potentially harmful for health. @edli11, finding similar results in another environment, suggest that an explanation for the larger social health inequalities in present-day society now is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking, smoking and more inclined to take risks.  Alcohol has been suggested as part of the large gender differences in the 19th century [@wil99; @edv92], and a more risk-taking life may be an important component of male mortality [@cour00]. Different historical contexts promote different ways to spend economic resources [@rasp06; @jaadla17].

@@

Overweight, obesity and physical inactivity were probably more common in higher social classes. @vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers, something they suggest can be explained by the higher tobacco consumption among men in these groups. The social pattern among women was however reversed in line with the present study. There is no solid information on smoking in different social classes before the 1960s, but there are international studies indicating higher prevalence and consumption of tobacco in higher classes during the early phases of the tobacco epidemic, illustrated for example by the fact that smoking initially was more prevalent among physicians. There was also a clear association between higher incomes and tobacco consumption [@loco94].
